\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-07 chapters/abstract.tex
Intro, P300 stimulation
cover page
Abstract

Main contribution: improve gaze-independent decoding by isolating component
related to visuospatial attention from visual components

>>>>

\chapter*{Abstract}
Individuals with severe motor impairments, such as those with \ac{lis}
or eye-motor dysfunction, face substantial challenges when using traditional
gaze-dependent brain-computer interfaces \acp{bci}.
These systems require users to direct their gaze toward specific targets, a
task that becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution of this work is the compensation for ERP latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

Beyond gaze independence, the work also advances general \ac{erp} decoding
techniques by refining the structure of linear and multilinear estimators.
These methods improve classification accuracy across a range of \ac{bci}
conditions, particularly when training data are limited.
The introduction of structured regularization in both linear and multilinear
models enhances the interpretability of the classifiers and reduces training
time and computational complexity.
This allows for more efficient training and better performance on unseen data,
contributing to more reliable systems that are adaptable to various contexts and user needs.

The proposed methods were validated in experiments involving both healthy
individuals and 7 individuals with severe physical impairment and impaired eye-motor control.
Data were recorded from a control group, as well as from seven individuals with
gaze impairments.
These experiments demonstrated the robustness of the novel decoding methods,
showing that the system could effectively decode covert attention even when
direct gaze was impossible.

\paragraph{Keywords:} brain-computer interface, gaze independence, covert
visuospatial attention, event-related potentials, linear decoding, multilinear
decoding, motor impairment, gaze impairment

### 2024-10-07 chapters/bci.tex
Abstract

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}
\todo{acro usage in chapters}

@@

\todomvh{"trial" could  be confused with intensification rounds e.g.}
\todo{remove trial alltogether}
Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial\footnote{Given that selections
are independent on each other. This formula requires some adaptations in the
case of sequential or hierarchical selection.}, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or
responses.
\end{enumerate*}

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation in a
    single channel.}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth, height=0.6180\textwidth]{example-image-b}
  \end{subfigure}
\end{figure}

\todo{four pane figure:
  photo of someone using a P300 bci
  full spatiotemporal ERP response
}

@@

\emph{The N1 component} peaks around 150 to 200 ms post-stimulus.
It represents an extension of sensory processing related to both gaze fixation
and attention.
For attended stimuli, the amplitude is greater, indicating prioritization for
further cognitive processing.
This highlights a more selective form of attention allocation, demonstrating
the impact of both types of attention on visual processing.

\emph{The P2 component} occurs between 200 to 300 ms.
It reflects higher-order processing, particularly in contexts demanding
stimulus evaluation and feature detection.
P2 is sensitive to attention.
When participants actively engage with specific features or categories of
stimuli, P2 shows increased amplitude.
Conversely, P2 may be less responsive when gaze is not directed toward the
stimulus but attention is still maintained.

\emph{The N2 component} peaking around 200 to 350 ms is associated with cognitive
control processes such as conflict monitoring and inhibition.
It is particularly pronounced in tasks requiring differentiation between
competing stimuli.
N2 reflects the allocation of attentional resources required to resolve
conflicts, irrespective of whether the gaze is directly on the competing
stimuli or not.
Higher amplitudes in N2 are seen when cognitive demands increase, indicating
the influence of attention strategies.

\emph{The P3 component} is elicited in oddball paradigms and peaks around 300 ms
post-stimulus.
It reflects attentional engagement and the processing of rare or unexpected
events.
The P3 is typically subdivided into two subcomponents: P3a and P3b.
The P3a is associated with the allocation of attention to novel or unexpected
stimuli, indicating the initial detection of a change in the environment.
It is particularly responsive to stimuli that capture attention, whether
through gaze fixation or attention.
P3a typically exhibits a frontal distribution and peaks earlier than P3b.

@@

\emph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
Sustained attention and emotional processing of stimuli are reflected in the
LPP.
Typically enhanced for emotionally salient or motivationally relevant
information, the LPP serves as an indicator of cognitive appraisal beyond mere
recognition tasks.
This component can be influenced by attention when the participant evaluates
the emotional significance of stimuli, even if gaze is not directed at them.

>>>>

\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}
\todo{acro usage in chapters}

@@

Finally, an important concept in the evaluation of a \ac{bci} decoder, and of \acp{bci}
in general, is \ac{itr}.
\Ac{itr} reflects how fluently a \ac{bci} can be used for communication and can
be calculated as
\begin{equation}
	\text{ITR} = Q\left(\log_2N+P\log_2P+(1-P)\log_2\frac{1-P}{N-1}\right)
\end{equation}
\Ac{itr} is expressed in bits/s and is dependent on $N$, the number of
different options that can be selected per trial\footnote{Given that selections
are independent on each other. This formula requires some adaptations in the
case of sequential or hierarchical selection.}, $P$, the selection accuracy
of the decoder, and $Q$, the number of trials per second.
The parameters $N$, $P$ and $Q$ of this formula give us some insight in the
building blocks a successful, high-\ac{itr} \ac{bci}.
To improve performance, we can aim to
\begin{enumerate*}[label={\arabic*})]
\item increase $N$ by selecting a paradigm and interface that offer a
broad range of informative selections per trial,
\item increase $P$ by engineering more performant machine learning methods for
decoding, and
\item increase $Q$ by selecting a paradigm that allows fast stimulation or
responses.
\end{enumerate*}

@@

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation in a
    single channel.}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth, height=0.6180\textwidth]{example-image-b}
  \end{subfigure}
\end{figure}
\todo{full spatiotemporal erp response}

@@

\paragraph{The N1 component} peaks around 150 to 200 ms post-stimulus.
It represents an extension of sensory processing related to both gaze fixation
and attention.
For attended stimuli, the amplitude is greater, indicating prioritization for
further cognitive processing.
This highlights a more selective form of attention allocation, demonstrating
the impact of both types of attention on visual processing.

\paragraph{The P2 component} occurs between 200 to 300 ms.
It reflects higher-order processing, particularly in contexts demanding
stimulus evaluation and feature detection.
P2 is sensitive to attention.
When participants actively engage with specific features or categories of
stimuli, P2 shows increased amplitude.
Conversely, P2 may be less responsive when gaze is not directed toward the
stimulus but attention is still maintained.

\paragraph{The N2 component} peaking around 200 to 350 ms is associated with cognitive
control processes such as conflict monitoring and inhibition.
It is particularly pronounced in tasks requiring differentiation between
competing stimuli.
N2 reflects the allocation of attentional resources required to resolve
conflicts, irrespective of whether the gaze is directly on the competing
stimuli or not.
Higher amplitudes in N2 are seen when cognitive demands increase, indicating
the influence of attention strategies.

\paragraph{The P3 component} is elicited in oddball paradigms and peaks around 300 ms
post-stimulus.
It reflects attentional engagement and the processing of rare or unexpected
events.
The P3 is typically subdivided into two subcomponents: P3a and P3b.
The P3a is associated with the allocation of attention to novel or unexpected
stimuli, indicating the initial detection of a change in the environment.
It is particularly responsive to stimuli that capture attention, whether
through gaze fixation or attention.
P3a typically exhibits a frontal distribution and peaks earlier than P3b.

@@

\paragraph{The late positive potential (LPP)} is a later component occurring beyond 400 ms.
Sustained attention and emotional processing of stimuli are reflected in the
LPP.
Typically enhanced for emotionally salient or motivationally relevant
information, the LPP serves as an indicator of cognitive appraisal beyond mere
recognition tasks.
This component can be influenced by attention when the participant evaluates
the emotional significance of stimuli, even if gaze is not directed at them.

### 2024-10-07 chapters/conclusion.tex
Abstract

\section{Current \& future work}
On the decoder development side, we have partially implemented some interesting
extensions of the proposed methods.
The Kronecker-structured beamformer can be generalized to the \ac{lda} case.
Furthermore, we obtained promising results when extending the single
Kronecker product covariance model to a sum of Kronecker product terms, since
the \ac{eeg} covariance is probably better expressed by such a
structure~\cite{Bijma2005}.

To overcome the limitations introduced by the presence of multiple \ac{erp}
components to latency estimation, we have also developed a multi-component
version of the \ac{wcble} algorithm.
This algorithm should be able to separate mixed \ac{erp} component clusters
based on their temporal coherence, in a similar maner to~\textcite{Ouyang2017}.
It could also yield favorable decoding results in a broader range of settings
than overt \ac{vsa} since it should be able to account for the presence of both
visual and attentional \ac{erp}components}.

@@

\section{Conclusion}

>>>>

\section{Current \& future work}
On the decoder development side, we have partially implemented some interesting
extensions of the proposed methods.
The Kronecker-structured beamformer can be generalized to the \ac{lda} case.
Furthermore, we obtained promising results when extending the single
Kronecker product covariance model to a sum of Kronecker product terms, since
the \ac{eeg} covariance is probably better expressed by such a
structure~\cite{Bijma2005}.
The Kronecker-lda approach has also be extended to the combined
space-time-frequency domain, where more information can be modeled at the cost
of increased dimensionality.
A properly structured covariance model can strongly regularize the problem.

To overcome the limitations introduced by the presence of multiple \ac{erp}
components to latency estimation, we have also developed a multi-component
version of the \ac{wcble} algorithm.
This algorithm should be able to separate mixed \ac{erp} component clusters
based on their temporal coherence, in a similar maner to~\textcite{Ouyang2017}.
It could also yield favorable decoding results in a broader range of settings
than overt \ac{vsa} since it should be able to account for the presence of both
visual and attentional \ac{erp}components.

@@
